---
title: "Modelling interventions different scales"
editor: visual
---
{{< include _ch4_src.qmd >}}
## Motivation

-   Reduce the total epidemic size/effects by intervening in strategic locations

-   Spatial resolution allows for precise targeting. Behavioural intervention (restriction of mixing) Mixing matrix is modified during simulation

## Formulation

-   Parameters

    -   Trigger: condition at which intervention occurs

    -   Effect:
      
        -   `local` reduce the intra-patch mixing

        -   `travel` reduce inter-patch mixing

        -   `total` reduce intra- and inter-patch mixing

## Implimentation

### local intervention

#### final size
```{r}
sim_stats %>% 
  ggplot(aes(x = R0, y = finalsize, col = Intervention), alpha = 0.4) +
  geom_point() +
  facet_grid(mu ~ SA)
```

#### peak size
```{r}
sim_stats %>% 
  ggplot(aes(x = R0, y = peaksize, col = Intervention)) +
  geom_point() +
  facet_grid(mu ~ SA)
```
#### duration
```{r}
sim_stats %>% 
  filter(duration < 1000, mu == "1.0") %>% 
  ggplot(aes(x = finalsize, y = duration, color= Intervention)) +
  geom_point()+
  facet_grid(R0 ~ SA)

sim_stats %>% 
  filter(duration < 1000, mu == "1.0") %>% 
  ggplot(aes(x = finalsize, y = duration, color= Intervention)) +
  geom_point()+
  facet_grid(SA ~ R0)
```
#choose single simulation and look at time evolution
    -look at patch counts 
    - patch counts
